Clinical Language Encoding with ModernBERT

### 2025-04-15 article/article.qmd
update

---
title: "Clinical Language Encoding with ModernBERT"
format:
  arxiv-pdf:
    keep-tex: true  
    linenumbers: false
    doublespacing: false
    runninghead: "A Preprint"
  arxiv-html: default
author:
  - name: Tyler M Cross
    affiliations:
      - name: University of California, Berkeley School of Information 
        department: Master's in Data Science
        address: 102 South Hall
        city: Berkeley, CA
        country: USA
        postal-code: 94720
    orcid: 0009-0003-3529-8222
    email: tyler.cross@berkeley.edu
    url: https://tylercross.me
  # - name: Someone Else
  #   affiliations:
  #     - name: State University of New York College of Environmental Science and Forestry
  #       department: Department of Sustainable Resources Management
  #       address: 1 Forestry Drive
  #       city: Syracuse, NY
  #       country: USA
  #       postal-code: 13210
abstract: |
  Medical coding—assigning standardized codes to clinical documentation—is a costly, labor-intensive process. Automating this task via multi-label classification faces challenges like long documents and extreme class imbalance. We present CLEM-ICD, replacing the RoBERTa/BERT base models common in prior frameworks with ModernBERT, leveraging its 8192-token context window to better process lengthy clinical notes. ModernBERT's optimized architecture (~138M parameters) enhances computational efficiency and simplifies implementation compared to prior approaches requiring complex modifications for shorter-context models. Evaluating on the MDACE dataset, CLEM-ICD achieves a Micro-F1 score of 47.6%, outperforming recent results by @edin2024explainable (41.9% Micro-F1). On the standard MIMIC-III benchmark, CLEM-ICD achieves a notably high Macro-F1 of 16.5%, significantly outperforming the current state-of-the-art on uncommon classes. CLEM-ICD demonstrates how modern architectural advancements, particularly large context windows applied to encoder-only models, can yield strong performance on complex tasks like automated medical coding. We release our code and models to foster further research.
keywords: 
  - clinical natural language processing
  - ModernBERT
  - MIMIC-IV
  - MIMIC-III
  - MDACE
  - multi-label classification
bibliography: bibliography.bib  
---

>>>>

---
title: "Clinical Language Encoding with ModernBERT"
format:
  arxiv-pdf:
    keep-tex: true  
    linenumbers: false
    doublespacing: false
    runninghead: "A Preprint"
  arxiv-html: default
author:
  - name: Tyler M Cross
    affiliations:
      - name: University of California, Berkeley School of Information 
        department: Master's in Data Science
        address: 102 South Hall
        city: Berkeley, CA
        country: USA
        postal-code: 94720
    orcid: 0009-0003-3529-8222
    email: tyler.cross@berkeley.edu
    url: https://tylercross.me
  # - name: Someone Else
  #   affiliations:
  #     - name: State University of New York College of Environmental Science and Forestry
  #       department: Department of Sustainable Resources Management
  #       address: 1 Forestry Drive
  #       city: Syracuse, NY
  #       country: USA
  #       postal-code: 13210
abstract: |
  Medical coding—assigning standardized codes to clinical documentation—is a costly, labor-intensive process. Automating this task via multi-label classification faces challenges like long documents and extreme class imbalance. We present CLEM-ICD, replacing the RoBERTa/BERT base models common in prior frameworks with ModernBERT, leveraging its 8192-token context window to better process lengthy clinical notes. ModernBERT's optimized architecture (~138M parameters) enhances computational efficiency and simplifies implementation compared to prior approaches requiring complex modifications for shorter-context models. Evaluating on the MDACE dataset, CLEM-ICD achieves a Micro-F1 score of 47.6%, outperforming recent results by @edin2024explainable (41.9% Micro-F1). On the standard MIMIC-III benchmark, CLEM-ICD achieves a notably high Macro-F1 of 16.5%, significantly outperforming the current state-of-the-art on uncommon classes. CLEM-ICD demonstrates how modern architectural advancements, particularly large context windows applied to encoder-only models, can yield strong performance on complex tasks like automated medical coding. We release our code and models to foster further research.
keywords: 
  - clinical natural language processing
  - ModernBERT
  - MIMIC-III
  - MDACE
  - multi-label classification
bibliography: bibliography.bib  
---

### 2025-04-15 article/article.tex
update

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Clinical Language Encoding with ModernBERT},
  pdfauthor={Tyler M Cross},
  pdfkeywords={clinical natural language
processing, ModernBERT, MIMIC-IV, MIMIC-III, MDACE, multi-label
classification},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{A Preprint }
\title{Clinical Language Encoding with ModernBERT}
\def\asep{\\\\\\ } % default: all authors on same column
\author{\textbf{Tyler M Cross}~\orcidlink{0009-0003-3529-8222}\\Master's
in Data Science\\University of California, Berkeley School of
Information\\Berkeley,
CA,\ 94720\\\href{mailto:tyler.cross@berkeley.edu}{tyler.cross@berkeley.edu}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Medical coding---assigning standardized codes to clinical
documentation---is a costly, labor-intensive process. Automating this
task via multi-label classification faces challenges like long documents
and extreme class imbalance. We present CLEM-ICD, replacing the
RoBERTa/BERT base models common in prior frameworks with ModernBERT,
leveraging its 8192-token context window to better process lengthy
clinical notes. ModernBERT's optimized architecture (\textasciitilde138M
parameters) enhances computational efficiency and simplifies
implementation compared to prior approaches requiring complex
modifications for shorter-context models. Evaluating on the MDACE
dataset, CLEM-ICD achieves a Micro-F1 score of 47.6\%, outperforming
recent results by Edin et al. (2024) (41.9\% Micro-F1). On the standard
MIMIC-III benchmark, CLEM-ICD achieves a notably high Macro-F1 of
16.5\%, significantly outperforming the current state-of-the-art on
uncommon classes. CLEM-ICD demonstrates how modern architectural
advancements, particularly large context windows applied to encoder-only
models, can yield strong performance on complex tasks like automated
medical coding. We release our code and models to foster further
research.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
clinical natural language
processing \sep ModernBERT \sep MIMIC-IV \sep MIMIC-III \sep MDACE \sep 
multi-label classification

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Clinical Language Encoding with ModernBERT},
  pdfauthor={Tyler M Cross},
  pdfkeywords={clinical natural language
processing, ModernBERT, MIMIC-III, MDACE, multi-label classification},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{A Preprint }
\title{Clinical Language Encoding with ModernBERT}
\def\asep{\\\\\\ } % default: all authors on same column
\author{\textbf{Tyler M Cross}~\orcidlink{0009-0003-3529-8222}\\Master's
in Data Science\\University of California, Berkeley School of
Information\\Berkeley,
CA,\ 94720\\\href{mailto:tyler.cross@berkeley.edu}{tyler.cross@berkeley.edu}}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Medical coding---assigning standardized codes to clinical
documentation---is a costly, labor-intensive process. Automating this
task via multi-label classification faces challenges like long documents
and extreme class imbalance. We present CLEM-ICD, replacing the
RoBERTa/BERT base models common in prior frameworks with ModernBERT,
leveraging its 8192-token context window to better process lengthy
clinical notes. ModernBERT's optimized architecture (\textasciitilde138M
parameters) enhances computational efficiency and simplifies
implementation compared to prior approaches requiring complex
modifications for shorter-context models. Evaluating on the MDACE
dataset, CLEM-ICD achieves a Micro-F1 score of 47.6\%, outperforming
recent results by Edin et al. (2024) (41.9\% Micro-F1). On the standard
MIMIC-III benchmark, CLEM-ICD achieves a notably high Macro-F1 of
16.5\%, significantly outperforming the current state-of-the-art on
uncommon classes. CLEM-ICD demonstrates how modern architectural
advancements, particularly large context windows applied to encoder-only
models, can yield strong performance on complex tasks like automated
medical coding. We release our code and models to foster further
research.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
clinical natural language
processing \sep ModernBERT \sep MIMIC-III \sep MDACE \sep 
multi-label classification
